Combinator Parsers: a Monad Transformer Approach

### 2013-05-01 article.tex
continue write-up of basic parser

\section{Parser Combinators}
Parser combinators have long been used for parsing in the 
functional programming community, represent a powerful and simple solution.
Their main advantages are ability to deal with complicated formats (see
Chomsky levels), language integration, and the declarativeness and
succinctness of the resulting parsers.
Parser combinator libraries typically include a set of basic parsers and a
set of combinators for combining little parsers into big ones.  Parsers are
typically functions whose type is something like:
\begin{verbatim}
[t] -> m ([t], a)
\end{verbatim}
where t is the token type, a is the result type, and m represents some
computational effects.
A very simple parser succeeds, consuming a single token, when the token stream
is not empty, and fails otherwise.  We can make the substitutions t -> Char,
m -> Maybe, and a -> Char and implement the function:
\begin{verbatim}
    item :: [Char] -> Maybe ([Char], Char)
    item (t:ts) = Just (ts, t)
    item []     = Nothing

    -- examples
    ghci$ item "abcde"
    Just ("bcde", 'a')

@@

We can also come up with a simple combinator that takes a parser and a function
as input, and succeeds if the parser succeeds and the function applied to the
parser's result succeeds:
\begin{verbatim}
    check :: (a -> Bool) -> ([Char] -> Maybe ([Char], a)) -> [Char] -> Maybe ([Char], a)
    check f p = p >>= \(ts, x) -> if (f x) then Just (ts, x) else Nothing

    -- examples
    ghci$ check (== 'a') item "abcde"
    Just ("bcde", 'a')

@@

\section{Monad transformers}

\section{Introducing Woof:  a Simple Lisp}

The following examples will all use Woof, a simple dialect of Lisp,
progressively adding features to a simplistic initial implementation
to create a usable parser.

@@

\section{Example 1: recognizing}

\subsection{An Elementary Parser}
The basic types of parsers will be something like
\begin{verbatim}[t] -> m ([t], a)\end{verbatim}, or, a function that consumes
part of a list of tokens of type t with computational effects m and
result a.
A very basic and fundamental parser is the one that succeeds whenever
the list of tokens is not empty, consuming one token and presenting it
as the result, and fails when the list is empty.  We'll call this item,
and we can easily implement it in terms of the MonadState type class:
\begin{verbatim}
item :: (MonadState [t] m, Plus m) => m t
item =
    get >>= \xs -> case xs of
                        (t:ts) -> put ts *> pure t;
                        []     -> zero;
\end{verbatim}

\subsection{Token Parsers}
Let's start with parsers for recognizing our most basic syntactic elements:  tokens.
These include the seven listed above, as well as whitespace and comments.

>>>>

\section{Parser Combinators}
Parser combinators have long been used for parsing in the 
functional programming community, represent a powerful and simple solution.
Their main advantages are ability to deal with complicated formats (see
Chomsky levels), language integration, and the declarativeness and
succinctness of the resulting parsers.
Parser combinator libraries typically include a set of basic parsers and a
set of combinators for combining little parsers into big ones.  Parsers are
typically functions whose type is something like:
\begin{verbatim}
[t] -> m (a, [t])
\end{verbatim}
where t is the token type, a is the result type, and m represents some
computational effects.
A very simple parser succeeds, consuming a single token, when the token stream
is not empty, and fails otherwise.  We can make the substitutions t -> Char,
m -> Maybe, and a -> Char and implement the function:
\begin{verbatim}
    item :: [Char] -> Maybe (Char, [Char])
    item (t:ts) = Just (t, ts)
    item []     = Nothing

    -- examples
    ghci$ item "abcde"
    Just ('a', "bcde")

@@

We can also come up with a simple combinator that takes a parser and a function
as input, and succeeds if the parser succeeds and the function applied to the
parser's result succeeds:
\begin{verbatim}
    check :: (a -> Bool) -> ([Char] -> Maybe (a, [Char])) -> [Char] -> Maybe (a, [Char])
    check f p = p >>= \(x, ts) -> if (f x) then Just (x, ts) else Nothing

    -- examples
    ghci$ check (== 'a') item "abcde"
    Just ('a', "bcde")

@@

\section{Monad transformers}
Monad transformers are a technique for combining monads to form new monads.
The motivation behind them is that while single monads are quite useful, oftentimes
the semantics of multiple monads are required simultaneously; it is then desirable
to have a modular approach for creating combined monads.

There are multiple to approaches to combining monads and to monad transformers
themselves, but we'll follow that of the standard transformers/mtl library, since
it's easy to obtain, easy to use, and quite effective in practice.

Along the way, we'll also explore some of the consequences of combining monads,
both with respect to semantics and to the definition of the Haskell language.  We'll
also bump in to some of the limitations, trade-offs and design decisions of the 
transformer/mtl library.

\section{Introducing Woof:  a Simple Lisp}
The simple language we'll use as a motivation for building parsers is 
Woof, a simple dialect of Lisp.  We'll be 
progressively adding features to a simplistic initial implementation
on our way to creating a practical, feature-rich and usable parser.

@@

\section{Example 1: Recognition and Tree-Building}
The first parser we build will be responsible for determining whether input
text conforms to the language definition and simultaneously build an 
Abstract Syntax Tree representing of the structure of the recognized input.

Here's the AST definition we'll use:
\begin{verbatim}
data AST
    = ANumber Integer
    | ASymbol String
    | AString String
    | ALambda [String] [AST]
    | ADefine (Maybe String) String AST
    | AApp    AST  [AST]
  deriving (Show, Eq)
\end{verbatim}
Plus we'll need a type for our parsers.  We'll use the basic type mentioned 
above -- [t] -> m (a, [t]) -- as our starting point.  As luck would have it, that
corresponds to our first monad transformer:  StateT (Control.Monad.State), so we
can write:
\begin{verbatim}
type Parser a = StateT [Char] Maybe a
\end{verbatim}
Which means that our Parser is a State monad transformer, with the Maybe type
as its underlying Monad parameter.

Now let's get down to the business of building parsers, starting with parsers
for our most basic syntactic elements:  tokens.
These include the four braces, symbols, strings, numbers, whitespace, and comments.
To recognize braces, we need to re-implement the item and check combinators that
we did before.  Fortunately, using the StateT transformer makes it simpler
to do so:
\begin{verbatim}
item :: (MonadState [t] m, Plus m) => m t
item =
    get >>= \xs -> case xs of
                        (t:ts) -> put ts *> pure t;
                        []     -> zero;
\end{verbatim}
Although we now also need to discuss the Plus type class, why we need it instead of
Alternative, and what its instances are.
